(Interacting with)? \\ Modal Logics in the Coq Proof Assistant

### 2014-12-10 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
modified draft paper

\title{
Modal Logics in the Coq Proof Assistant
}

@@

\institute{
  Dahlem Center for Intelligent Systems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
}

@@

\begin{abstract}  
This paper describes an embedding of higher-order
modal logics in the \Coq proof assistant. {\Coq}'s capabilities are
thus extended in a minimalistic manner, which is nevertheless
sufficient for the formalization of significant modal proofs. The
elegance, flexibility and convenience of this approach, from a user
perspective, are illustrated here with the successful formalization of
Gödel's ontological argument.  
\end{abstract}

@@

Modal logics extend usual formal logic languages by adding modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, although $A \imp \Box A$ is not necessarily a
theorem. Various informal concepts, such as \emph{necessity and
possibility}, \emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, have been formalized with the help of
modal operators. Therefore, modal logics \cite{ModalLogic} are well-
established for a wide variety of application domains, ranging from
philosophy \cite{ToDo} to formal software verification
\cite{ToDo:TLAPlus}.

@@

Another recently explored possibility \cite{J23,B9,J22} is the
embedding of first-order and even higher-order modal logics into classical
higher-order logics, for which existing higher-order automated theorem
provers \cite{ToDo:Leo,Satallax} exist.  Embedding is flexible,
because various modal logics can be easily supported by adding their
characteristic axioms.  The embedding approach can also be easily
adapted to support multiple modalities, and it is possible to replace
constant domain quantifiers by varying or cumulative domain
quantifiers.  Moreover, the approach is relatively simple to
implement, because it does not require any modification in the source
code of the higher-order prover. The prover can be used as is, and
only the input files provided to the prover must be especially
encoded. Furthermore, the efficacy and efficiency of the embedding
approach has been confirmed in philosophically interesting and
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

@@

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{Scott} of Gödel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{ToDo}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with first-
order theorem provers by Oppenheimer and Zalta \cite{Zalta}. Gödel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{ArXiv,AFP}. But this
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant.

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper.} (e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

\begin{calculus}
{An arbitrary natural deduction calculus}
{fig:NaturalDeduction}
\begin{small}

@@

Since the modal operators are, in our embedding, essentially just
abbreviations for quantifiers guarded by reachability conditions, the
typical tactics for quantifiers can be used, in principle.  However,
this exposes the user to the technicalities of the embedding,
requiring him to deal with possible worlds and their reachability
explicitly. In order to obtain transparency also for the modal
operators, we have implemented a few specialized tactics using \Coq's
Ltac language.

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a
modern culmination of this tradition, following particularly the
footsteps of Leibniz. Various slightly different versions of axioms
and definitions have been considered by G\"{o}del and by several
philosophers who commented on his proof (cf. 
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow all the
steps in Scott's notes. Scott's assertions are emphasized with
comments. In contrast to the formalization in \Isabelle \cite{AFP},
where automation via \texttt{Metis} \cite{Hurd03first-orderproof} and
\texttt{Sledgehammer} \cite{Sledgehammer} using tools such as
\texttt{Nitpick} \cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and
\texttt{Satallax} \cite{Satallax} has been successfully employed, the
formalization in \Coq used no automation. This was a deliberate
choice, mainly because it allowed a qualitative evaluation of the
convenience of the possible world semantic embedding approach for
\emph{interactive} theorem proving. Moreover, in order to formalize
precisely Scott's version and not another automatically found version,
automation would have to be heavily limited anyway. Furthermore, the
deliberate preference for simple tactics (mostly \emph{intro},
\emph{apply} and the modal tactics described in Section
\ref{sec:Tactics}) results in proof scripts that closely correspond to
natural deduction proofs. This hopefully makes the formalization more
accessible to those who are not experts in \Coq's tactics but are
nevertheless interested in G\"odel's proof.

Gödel's proof requires \Coq's classical logic libraries as well as the
\texttt{Modal} library developed by us and described in sections
\ref{sec:Embedding} and \ref{sec:Tactics}.

@@

The successful formalization of Scott's version of  G\"odel's
ontological argument indicates that the \emph{embedding} of  modal
logics into higher-order logics via the \emph{possible world
semantics} is a viable approach for fully interactive theorem proving
within modal logics. Our lightweight implementation of the embedding
(described in sections \ref{sec:Embedding} and \ref{sec:Tactics})
takes special care to hide the underlying possible world machinery
from the user. An inspection of the proof scripts in Section
\ref{sec:proof} shows that this goal has been achieved. The user does
not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for
him/her. Moreover, for subgoals that do not involve modalities, the
user has all the usual interactive tactics at his/her disposal.

Although fully automated (as opposed to interactive) theorem proving
is beyond the scope of this paper, it is worth mentioning that all
lemmas and theorems in sections \ref{sec:Embedding},
\ref{sec:SimpleExamples} and \ref{sec:BeyondK} could have been proven
automatically using \Coq's \texttt{firstorder} tactic. The
implementation of hints to allow \Coq's automatic tactics to take full
advantage of the embedding and the modal axioms still remains for
future work.

The theological implications of the verified correctness of Gödel's
proof certainly depend on a critical discussion of the underlying
concepts, definitions and axioms, which is beyond the scope of this
paper.  Clearly, the application of theorem proving technology to the
domain of theoretical philosophy can --- as already pictured by
Leibniz --- be very fruitful for both areas. We hope that the
infrastructure  that we have implemented for interactive and automated
reasoning in higher-order modal logics will be useful outside
philosophy as well.

\paragraph{Acknowledgements:} we thank Cedric Auger and Laurent Théry, 
for their answers to our questions about Ltac in the Coq-Club mailing-list.

@@

>>>>

\title{
(Interacting with)? \\ Modal Logics in the Coq Proof Assistant
}

@@

\institute{
  % Dept. of Mathematics and Computer Science, 
  Freie Universit\"{a}t Berlin, Germany,  %\\
  \email{c.benzmueller@fu-berlin.de}
  \and 
  % Theory and Logic Group, 
  Vienna University of Technology, Austria, %\\
   \email{bruno@logic.at}
}

@@

\begin{abstract}  
  This paper describes an embedding of higher-order modal logics in
  the \Coq proof assistant. {\Coq}'s capabilities are extended in a
  minimalistic manner, which is nevertheless sufficient for the
  formalization of significant, non-trivial modal logic proofs. The
  elegance, flexibility and convenience of this approach, from a user
  perspective, are illustrated here with the successful formalization
  of Gödel's ontological argument.
\end{abstract}

@@

Modal logics extend usual formal logic languages by adding modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, although $A \imp \Box A$ is not necessarily a
theorem. Various informal concepts, such as \emph{necessity and
possibility}, \emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, have been formalized with the help of
modal operators. Therefore, modal logics \cite{ModalLogic} are well-
established for a wide variety of application domains, ranging from
philosophy \cite{ToDo} to formal software verification
\cite{ToDo:TLAPlus}.\footnote{ChrisToBruno: \cite{ModalLogic} should
  suffice as reference, applications are addressed in Part 4.}

@@

Another recently explored possibility \cite{J23,B9,J22} is the
embedding of first-order and even higher-order modal logics into classical
higher-order logics, for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist.  Embedding is flexible,
because various modal logics can be easily supported by adding their
characteristic axioms.  The embedding approach can also be easily
adapted to support multiple modalities, and it is possible to replace
constant domain quantifiers by varying or cumulative domain
quantifiers.  Moreover, the approach is relatively simple to
implement, because it does not require any modification in the source
code of the higher-order prover. The prover can be used as is, and
only the input files provided to the prover must be especially
encoded. Furthermore, the efficacy and efficiency of the embedding
approach has been confirmed in philosophically interesting and
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

@@

Propositional and quantified modal logics have applications in various
fields, including, for example, verification, agent technologies,
artificial intelligence, computational linguistics, and computational
philosophy. Our contribution may thus serve as a starting point for
many interesting further projects.

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{ScottNotes} of Gödel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{ToDo}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with first-
order theorem provers by Oppenheimer and Zalta \cite{Zalta}. Gödel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{arXiv,AFP}. This
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant. Hence we here reveal the proof
structure on a cognitively adequate level of
detail. Such detailed proof structure has been hidden in our
other papers on the subject due to the use of automated theorem
provers. In comparison to \cite{Rushby,Zalta}, however, our
contribution stands out with its surprising technical 
simplicity and elegance.

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper.} \footnote{ChrisToBruno: We need to briefly address here
    the fact that this calculus is classical while Coq pure is
    constructive; one or two sentences should be fine.} 
(e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

\begin{calculus}
{A (classical)  The natural deduction calculus}
{fig:NaturalDeduction}
\begin{small}

@@

Since the modal operators are, in our embedding, essentially just
abbreviations for quantifiers guarded by reachability conditions, the
typical tactics for quantifiers can be used, in principle.  However,
this exposes the user to the technicalities of the embedding,
requiring him to deal with possible worlds and their reachability
explicitly. In order to obtain transparency also for the modal
operators, we have implemented a few specialized tactics using \Coq's
Ltac language. We present these tactics next.

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern
culmination of this tradition, following particularly the footsteps of
Leibniz. Various slightly different versions of axioms and definitions
have been considered by G\"{o}del and by several philosophers who
commented on his proof (cf.
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this Section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow precisely
all the steps in Scott's notes. Scott's assertions are emphasized
below with comments. In contrast to the formalization in \Isabelle
\cite{AFP}, where automation via \texttt{Metis}
\cite{Hurd03first-orderproof} and \texttt{Sledgehammer}
\cite{Sledgehammer} using tools such as \texttt{Nitpick}
\cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and \texttt{Satallax}
\cite{Satallax} has been successfully employed, the formalization in
\Coq used no automation. This was a deliberate choice, mainly because
it allowed a qualitative evaluation of the convenience of the
embedding approach for \emph{interactive} theorem proving. Moreover,
in order to formalize precisely Scott's version and not another
automatically found version, automation would have to be heavily
limited anyway. Furthermore, the deliberate preference for simple
tactics (mostly \emph{intro}, \emph{apply} and the modal tactics
described in Section \ref{sec:Tactics}) results in proof scripts that
closely correspond to natural deduction proofs. This hopefully makes
the formalization more accessible to those who are not experts in
\Coq's tactics but are nevertheless interested in G\"odel's proof.

Gödel's proof requires \Coq's classical logic libraries as well as the
\texttt{Modal} library developed by us and described in Sections
\ref{sec:Embedding} and \ref{sec:Tactics}.

@@

The successful formalization of Scott's version of  G\"odel's
ontological argument indicates that the \emph{embedding} of
higher-order modal
logics into higher-order logics via the \emph{possible world
semantics} is a viable approach for fully interactive theorem proving
within (even non-trivial) modal logics. Our lightweight implementation of the embedding
(described in Sections \ref{sec:Embedding} and \ref{sec:Tactics})
takes special care to hide the underlying possible world machinery
from the user. An inspection of the proof scripts in Section
\ref{sec:proof} shows that this goal has been achieved. The user does
not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for
him/her. Moreover, for subgoals that do not involve modalities, the
user has all the usual interactive tactics at his/her disposal.

Although fully automated (as opposed to interactive) theorem proving
is beyond the scope of this paper, it is worth mentioning that all
lemmas and theorems in Sections \ref{sec:Embedding},
\ref{sec:SimpleExamples} and \ref{sec:BeyondK} could have been proven
automatically using \Coq's \texttt{firstorder} tactic. The
implementation of hints to allow \Coq's automatic tactics to take full
advantage of the embedding and the modal axioms still remains for
future work.

The theological implications of the verified correctness of Gödel's
proof certainly depend on a critical discussion of the underlying
concepts, definitions and axioms, which is beyond the scope of this
paper.  Clearly, the application of theorem proving technology to the
domain of theoretical philosophy can --- as already pictured by
Leibniz --- be very fruitful for both areas.

The infrastructure that we have implemented for interactive and
automated reasoning in higher-order modal logics will clearly be
useful also outside philosophy; the range of potential applications is
in fact 
very wide.

\paragraph{Acknowledgements:} We thank Cedric Auger and Laurent Théry, 
for their answers to our questions about Ltac in the Coq-Club mailing-list.

@@

\bibitem{JensOtten}
C.~Benzm{\"u}lller, J. Otten, and Th. Raths.
\newblock Implementing and Evaluating Provers for First-order Modal
Logics. 
\newblock In {\em ECAI 2012}, pages 163-168. IOS Press, 2012.
